A weakly structured stem for human origins in Africa

### 2022-09-02 paper.tex
Molinaro reference

\subsection*{A Late Middle Stone Age common ancestry for contemporary humans}

@@

We quantify migration rates among populations after their divergence $\approx120ka$.
Migration, prior to the Bantu expansion, between the ancestors of the Nama and other groups  
is an order of magnitude weaker than that observed between between western and eastern Africans (Table \ref{tab:migration-rates}).
All models infer relatively high gene flow between Ethiopia and western Africa
($m\approx2\times10^{-4},$ the constant proportion of immigrant lineages per generation over 60ka).
We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. 
We observe significant gene flow from the Amhara and Oromo into the Nama, a signal which is likely a proxy for the movement of
eastern African caprid and cattle pastoralists \citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. While this gene flow is not apparent from the ADMIXTURE plot (Figure~\ref{fig:diversity}), 
the ancestry is likely grouped into the Khoe-San component which has drifted appreciably from its ancestral Eastern African source. 
Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
(Figure~\ref{fig:diversity}).

@@

To account for population structure prior to 135ka, three of our four models allowed for two or more
``stem'' populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase, as illustrated in Figure \ref{fig:mergers}:
1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or ``merger'') events from the
other stem population, so that recent populations are formed by mergers
of multiple ancestral populations. Depending on parameter values,
this scenario encompasses archaic hominin introgression and
fragmentation-and-coalescence models (such as Figure~\ref{fig:proposed-models}B and C).
For many parameters, confidence intervals based on bootstrapping are relatively narrow
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. Model assumptions
have, however, a larger impact on parameter estimates (and thus, real uncertainty).
To convey model uncertainty, we highlight features of the two inferred models
with high likelihoods. These are referred to as the
``multiple-merger'' and the ``continuous-migration'' models. Both allow for
migration between stem branches, but differ primarily in the timing of the
early divergence of stem populations and their relative $N_e$
(Figure~\ref{fig:best-fit-models}). The two models also differ in the mode of divergence
during the Middle Stone Age.

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained (standard error (SE) $=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
479ka. The timing of this divergence was also poorly constrained (SE $=166$ka).
These populations evolve independently until 119ka
(SE $=28$ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the western Africans and
eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25ka (SE $=0.6$ka).
The later Stem 2 contribution to the western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
Central Africa, broadly speaking. The differing proportions in the Nama and
eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

@@

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred
models. Many previous studies have found a reduction of coalescence rates
between $1000$ka and $100$ka in humans, and thus inferred an increase in $N_e$
during the same period \citep{Li2011-le}.  This increase in inferred $N_e$
could be due to either an increase in population size or to ancestral
population structure during the Middle Stone Age \citep{Mazet2016-wn}.  All
models, including the single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between 100ka-1Ma extending deeper in the past than the
Relate estimate (Figure~\ref{fig:supp-iicr-sim} and SI
Section~\ref{inferred_relate}). Whereas the single-origin model achieves this by an
increase in $N_e$ during that period, the best-fit model recapitulate this
pattern without corresponding population size changes.

@@

\subsection*{The Middle Stone Age in Africa}

By contrast, our inferred models paint a more consistent picture of the late
Middle Stone Age as a critical period of change, assuming that estimates from
the recombination clock accurately relate to geological chronologies
(SI Section~\ref{sec:supp-mutation}).
During the Middle Stone Age, the multiple merger model indicates three
major stem lineages in Africa, tentatively assigned to southern (Stem 1S),
eastern (Stem 1E) and western/central Africa (Stem 2). While the length of
isolation among the stems is variable across model fits, models with a period of 
divergence, isolation and then a merger event (i.e. a ``reticulation'')
out-performed models with bifurcating divergence and continuous gene flow.

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) 
from southern Africa moved northward thereby encountering Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
contemporary groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population $\approx120$ka.
The second event $\approx100$ka between Stem 1E and Stem 2,
results in the formation of the
ancestors of East/West Africans as well as later ``Out of Africans''. The rapid
rise in sea levels and increased precipitation during MIS 5e, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested, e.g., for the
Paleo-Agulhas plain \citep{Marean2014-pg}.

>>>>

\subsection*{A Late Pleistocene common ancestry for contemporary humans}

@@

We quantify migration rates among populations after their divergence $\approx120ka$.
Migration, prior to the Bantu expansion, between the ancestors of the Nama and other groups
is an order of magnitude weaker than that observed between between western and eastern Africans (Table \ref{tab:migration-rates}).
All models infer relatively high gene flow between Ethiopia and western Africa
($m\approx2\times10^{-4},$ the constant proportion of immigrant lineages per generation over 60ka).
We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations \citep{Molinaro2019-rf},
comprising over half of their genetic ancestry, estimated to be 64--65\%.
We observe significant gene flow from the Amhara and Oromo into the Nama, a signal which is likely a proxy for the movement of
eastern African caprid and cattle pastoralists \citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. While this gene flow is not apparent from the ADMIXTURE plot (Figure~\ref{fig:diversity}),
the ancestry is likely grouped into the Khoe-San component which has drifted appreciably from its ancestral eastern African source.
Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
(Figure~\ref{fig:diversity}).

@@

To account for population structure prior to 135ka, three of our four models allowed for two or more
``stem'' populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase, as illustrated in Figure \ref{fig:mergers}:
1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or ``merger'') events from the
other stem population, so that recent populations are formed by mergers
of multiple ancestral populations. Depending on parameter values,
this scenario encompasses archaic hominin introgression and
fragmentation-and-coalescence models (such as Figure~\ref{fig:proposed-models}B and C).
For many parameters, confidence intervals based on bootstrapping are relatively narrow
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. Model assumptions
have, however, a larger impact on parameter estimates (and thus, real uncertainty).
To convey model uncertainty, we highlight features of the two inferred models
with high likelihoods. These are referred to as the
``multiple-merger'' and the ``continuous-migration'' models. Both allow for
migration between stem branches, but differ primarily in the timing of the
early divergence of stem populations and their relative $N_e$
(Figure~\ref{fig:best-fit-models}). The two models also differ in the mode of divergence, with the multiple-merger featuring a population reticulation during the Middle Pleistocene.

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained (standard error (SE) $=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
479ka. The timing of this divergence was also poorly constrained (SE $=166$ka).
These populations evolve independently until 119ka
(SE $=28$ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the western Africans and
eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25ka (SE $=0.6$ka).
The later Stem 2 contribution to the western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
central Africa, broadly speaking. The differing proportions in the Nama and
eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

@@

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred
models. Many previous studies have found a reduction of coalescence rates
between $1000$ka and $100$ka in humans, and thus inferred an increase in $N_e$
during the same period \citep{Li2011-le}.  This increase in inferred $N_e$
could be due to either an increase in population size or to ancestral
population structure during the Middle Pleistocene \citep{Mazet2016-wn}.  All
models, including the single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between 100ka-1Ma extending deeper in the past than the
Relate estimate (Figure~\ref{fig:supp-iicr-sim} and SI
Section~\ref{inferred_relate}). Whereas the single-origin model achieves this by an
increase in $N_e$ during that period, the best-fit model recapitulate this
pattern without corresponding population size changes.

@@

\subsection*{The Formation of Population Structure in Africa}

By contrast, our inferred models paint a more consistent picture of the Middle to Late Pleistocene as a critical period of change, assuming that estimates from
the recombination clock accurately relate to geological chronologies
(SI Section~\ref{sec:supp-mutation}).
During the late Middle Pleistocene, the multiple merger model indicates three
major stem lineages in Africa, tentatively assigned to southern (Stem 1S),
eastern (Stem 1E) and western/central Africa (Stem 2). While the length of
isolation among the stems is variable across model fits, models with a period of 
divergence, isolation and then a merger event (i.e. a ``reticulation'')
out-performed models with bifurcating divergence and continuous gene flow.

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) 
from southern Africa moved northward thereby encountering Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
contemporary groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population $\approx120$ka.
The second event $\approx100$ka between Stem 1E and Stem 2,
results in the formation of the
ancestors of East/West Africans including the ancestors of ``Out of Africans''. The rapid
rise in sea levels and increased precipitation during MIS 5e interglacial, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested, e.g., for the
Paleo-Agulhas plain \citep{Marean2014-pg}.
